Alien species

### 2023-03-20 DRAFT_alien_species.Rmd
start to validate

Importing and sub-setting the main data file, fix duplicate _hovedøkosystem_, calculate area, split one column in two, make numeric, and select target variables:
```{r, message=F}
naturetypes <- sf::st_read(dsn = path) %>%
  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
  mutate(hovedøkosystem = recode(hovedøkosystem,
                                 "naturligÅpneOmråderUnderSkoggrensa" = "naturligÅpneOmråderILavlandet"),
         area = st_area(.)) %>%
  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  filter(NiN_variable_code %in% myVars)
```

```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator.", fig.height=8, fig.width=4}
ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
  geom_bar()+
  coord_flip()+
  theme_bw(base_size = 10)+
  theme(legend.position = "top",
        legend.title = element_blank(),
        legend.direction = "vertical")+
  guides(fill = "none")+
  xlab("")+
  ylab("Number of localities")+
  facet_wrap(.~hovedøkosystem, ncol=1, scale="free")
```

@@

```{r, message=F}
ano_eco <- c("vaatmark", "naturlig_apne", "semi_naturlig_mark")
ano <- sf::st_read(paste0(pData, "Naturovervaking_eksport.gdb"),
                   layer = "ANO_SurveyPoint") %>%
  dplyr::filter(hovedoekosystem_punkt %in% ano_eco)
```

@@

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")
```

There are som multisurfaces in here that I will convert to multoipolygons before plotting.

@@

I need to do a little trick and reverse the upper and lower reference
values for the normalisation to work. This is a bug which can be fixed
inside eaTools (see [GitHub issue](https://github.com/NINAnor/eaTools/issues/15)).

@@

This normalisation seems reasonable to me. I can save it as indicator
values. There is no point yet making this a time series, and I will
assign all the indicator value to the average year of the data.

@@

```{r}
st_crs(HIA) == st_crs(regions)
```

Since the HIA map and the _region_ map are completely overlapping (HIA was masked using the region map), we can get the
intersections

```{r, eval=F}
HIA_regions <- eaTools::ea_homogeneous_area(HIA,
                             regions,
                             keep1 = "infrastructureIndex",
                             keep2 = "region")
saveRDS(HIA_regions, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
```

```{r, include=F}
# silently read cached file
HIA_regions <- readRDS(paste0(pData, "cache/HIA_regions.rds"))
```

Create a new column by crossing region and HIF

```{r}
HIA_regions <- HIA_regions %>%
  mutate(region_HIF = paste(region, infrastructureIndex))
```

@@

I now have `r length(unique(HIA_regions$region_HIF))` unique areas (for each main ecosystem) that I will, given
there is data, extrapolate indicator values over.

@@

I will subset the `slitasje_data` into the three ecosystems. Note that
only for open wetland do we have good ecosystem delineation maps to base
a spatial averaging on.

```{r}
#Check spelling
#unique(slitasje_data$hovedøkosystem)

# Rename one level
slitasje_data <- slitasje_data %>%
  mutate(hovedøkosystem = recode(hovedøkosystem, "Våtmark" = "våtmark"))

#make valid
slitasje_data <- st_make_valid(slitasje_data)

#subset
wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "våtmark",]
seminat <- slitasje_data[slitasje_data$hovedøkosystem == "semi-naturligMark",]
natOpen <- slitasje_data[slitasje_data$hovedøkosystem == "naturligÅpneOmråderILavlandet",]
```

@@

```{r, eval=F}
wetland_stats <- ea_spread(indicator_data = wetlands,
          indicator = i_2020,
          regions = HIA_regions,
          groups = region_HIF,
          summarise = TRUE)

seminat_stats <- ea_spread(indicator_data = seminat,
          indicator = i_2020,
          regions = HIA_regions,
          groups = region_HIF,
          summarise = TRUE)

natOpen_stats <- ea_spread(indicator_data = natOpen,
          indicator = i_2020,
          regions = HIA_regions,
          groups = region_HIF,
          summarise = TRUE)

@@

all_stats <- rbind(wetland_stats,
                   seminat_stats,
                   natOpen_stats)
all_stats <- all_stats %>%
  separate(region_HIF,
           into = c("region", "HIF"),
           sep = " ")

saveRDS(all_stats, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
```

```{r, include=F}
#silently read cached file
all_stats <- readRDS(paste0(pData, "cache/all_stats.rds"))
```

```{r dense-stats, fig.height= 12, fig.width=6}
temp_plot <- all_stats %>%
  mutate(HIF_NA = case_when(
    n > 150 ~ HIF
  ))

myColour_list <- unique(temp_plot$myColour)
names(myColour_list) <-   myColour_list

ggarrange(
ggplot(temp_plot, aes(x = region, y = n, group = HIF, fill = HIF_NA))+
  geom_bar(position = "dodge", stat = "identity", col="black")+
  theme_bw(base_size = 20)+
  scale_fill_brewer(palette="Oranges")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
  coord_flip()+
  facet_wrap(.~eco),
ggplot(temp_plot, aes(x = region, y = w_mean, group = HIF, fill = HIF_NA))+
  geom_bar(position = "dodge", stat = "identity", col="black")+
  theme_bw(base_size = 20)+
  scale_fill_brewer(palette="Oranges")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
  coord_flip()+
  facet_wrap(.~eco),
ggplot(temp_plot, aes(x = region, y = sd, group = HIF, fill = HIF_NA))+
  geom_bar(position = "dodge", stat = "identity", col="black")+
  theme_bw(base_size = 20)+
  scale_fill_brewer(palette="Oranges")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
  coord_flip()+
  facet_wrap(.~eco),
ncol = 1)

@@

In the figure above, colors refer to Human Impact Factor (amount or frequency of infrastructure) 
and NA means that there are less than 150 data points which we set as a threshold for evaluating the indicator values.

In the top row
we see that the sample size (number of polygons) varies between
ecosystems. For naturally open ecosystems many polygons are from high
infrastructure areas, not surprisingly, and there are few points for the
lowest impact class (zero such polygons in *Sørlandet* for example). We
may therefore not be able to extrapolate indicator values for Naturally
open areas in in *Sørlandet* with low levels of infrastructure.

In semi-natural areas and wetlands, we have very little data from HIF
class 3.

We can also from the middle row see the association between the indicator
values (area weighted means) and the modified infrastructure index
(HIF). We should then not put any weight on the bars were there is
little data (transparent bars). Then we se the general pattern that the indicator 
value decreases with increasing HIF. But there are exceptions, like for Semi-natural areas
in Vestlandet.

Let us look at the indicator to pressure relationship across all ecosystems.

```{r, fig.cap="Indicator to pressure relationship across all ecosystems."}
temp_plot %>%
  filter(!is.na(HIF_NA)) %>%
  ggplot(aes(x = HIF_NA, y = w_mean))+
  geom_point(size=2, position = position_dodge2(.1))+
  geom_violin(alpha=0)+
  theme_bw()
```

The pattern is not perfect, and there is quite some noise. But this is perhaps also expected.

@@

```{r, fig.cap="Smaple size against indicator uncertainty."}
temp_plot %>%
  ggplot(aes(x = n, y = sd))+
  geom_point(size=2, position = position_dodge2(.1))+
  theme_bw()+
    facet_wrap(.~eco)
```

This shows that the uncertainty is inflated with sample sizes less than about 300-500.

```{r, fig.cap="Summary statistics for the slitasje indicator."}
DT::datatable(all_stats)
```

@@

```{r, eval=F}
corrCheck <- st_intersection(slitasje_data, HIA_regions)
saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
```

>>>>

Importing and sub-setting the main data file, fix duplicate _hovedøkosystem_, calculate area, split one column in two, make numeric, and select target variables:
```{r, message=F}
naturetypes <- sf::st_read(dsn = path) %>%
  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
  mutate(hovedøkosystem = case_match(hovedøkosystem,
                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
                                  .default = hovedøkosystem),
         area = st_area(.)) %>%
  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  filter(NiN_variable_code %in% myVars)
```

```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator.", fig.height=8, fig.width=6}
ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
  geom_bar()+
  coord_flip()+
  theme_bw(base_size = 10)+
  theme(legend.position = "top",
        legend.title = element_blank(),
        legend.direction = "vertical")+
  guides(fill = "none")+
  xlab("")+
  ylab("Number of localities")+
  facet_wrap(.~hovedøkosystem, ncol=1, scale="free")
```

@@

```{r, message=F}
ano_eco <- c("vaatmark", "naturlig_apne", "semi_naturlig_mark")
ano <- sf::st_read(paste0(pData, "Naturovervaking_eksport.gdb"),
                   layer = "ANO_SurveyPoint") %>%
  dplyr::filter(hovedoekosystem_punkt %in% ano_eco) %>%
   mutate(hovedoekosystem_punkt = case_match(hovedoekosystem_punkt,
                                       "vaatmark" ~ "våtmark",
                                       "naturlig_apne" ~ "naturligÅpneOmråderUnderSkoggrensa",
                                       "semi_naturlig_mark" ~ "semi-naturligMark",
                                       .default = hovedoekosystem_punkt))
```

@@

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")

```{r, include=F, eval=FALSE}
# If I import the manuicipality names:
muni_names <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "administrativenhetnavn")
#The encoding option fail for gdb-files, so I fix the names again here
muni_names$navn <- stringi::stri_encode(muni_names$navn, from = "UTF-8", to = "UTF-8")

# ... there is no common ID to match these with the municipality polygons!!
```

There are some multi-surfaces in here that I will convert to multi-polygons before plotting.

@@

@@

There is no point yet making this a time series, and I will
assign all the indicator value to the average year of the data.

@@

```{r}
st_crs(HIA) == st_crs(muni)
```

Then we get the intersections (unique combinations)

```{r, eval=F}
HIA_muni <- eaTools::ea_homogeneous_area(HIA,
                             muni,
                             keep1 = "infrastructureIndex",
                             keep2 = "kommunenummer")
saveRDS(HIA_muni, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_muni.rds")
```

```{r, include=F}
# silently read cached file
HIA_muni <- readRDS(paste0(pData, "cache/HIA_muni.rds"))
```

Create a new column by crossing municipality number and HIF

```{r}
HIA_muni <- HIA_muni %>%
  mutate(muni_HIF = paste("ID", kommunenummer, infrastructureIndex, sep="_"))
```

@@

I now have `r length(unique(HIA_muni$muni_HIF))` unique areas (for each main ecosystem) that I will, given
there is data, extrapolate indicator values over.

@@

I will subset the `alien_data` into the three ecosystems. Note that
only for open wetland do we have good ecosystem delineation maps to base
a spatial averaging on.

```{r}
#make valid
alien_data <- st_make_valid(alien_data)

#subset
wetlands <- alien_data[alien_data$hovedøkosystem == "våtmark",]
seminat <- alien_data[alien_data$hovedøkosystem == "semi-naturligMark",]
natOpen <- alien_data[alien_data$hovedøkosystem == "naturligÅpneOmråderUnderSkoggrensa",]
```

@@

```{r, eval=F}
wetland_stats <- ea_spread(indicator_data = wetlands,
          indicator = i_2020,
          regions = HIA_muni,
          groups = muni_HIF,
          summarise = TRUE)

seminat_stats <- ea_spread(indicator_data = seminat,
          indicator = i_2020,
          regions = HIA_muni,
          groups = muni_HIF,
          summarise = TRUE)

natOpen_stats <- ea_spread(indicator_data = natOpen,
          indicator = i_2020,
          regions = HIA_muni,
          groups = muni_HIF,
          summarise = TRUE)

@@

all_stats <- rbind(wetland_stats,
                   seminat_stats,
                   natOpen_stats)
all_stats <- all_stats %>%
  separate(muni_HIF,
           into = c("tempLink", "municipalityNumber", "HIF"),
           sep = "_")

#saveRDS(all_stats, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")
saveRDS(all_stats, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")

@@

```{r, include=F}
#silently read cached file
all_stats <- readRDS(paste0(pData, "cache/all_stats_alienSpecies.rds"))
```

We have data from `r length(unique(all_stats$municipalityNumber))` out of `r nrow(muni)` municipalities. This figure shows how many municipalities we have at least one data point for, specific for each human impact factor level:

```{r, fig.cap="Number of municipalities we have at least one data point for, specific for each human impact factor level."}
all_stats %>%
  group_by(HIF, eco) %>%
  summarise(n = n()) %>%
  ggplot()+
  geom_bar(aes(x = HIF, y = n), stat ="identity")+
  theme_bw()+
  ylab("Number of municipalities")+
  facet_wrap(.~eco)
```

If we say that we need 20 data points for each HIA and municipality combintaion in order to extrapolate to the entire HIA for that municipality, we get much less to work with:

```{r, fig.cap="Number of municipalities we have at least 20 data points for, specific for each human impact factor level."}
all_stats %>%
  filter(n>20) %>%
  group_by(HIF, eco) %>%
  summarise(n = n()) %>%
  ggplot()+
  geom_bar(aes(x = HIF, y = n), stat ="identity")+
  theme_bw()+
  ylab("Number of municipalities")+
  facet_wrap(.~eco)

```{r, fig.cap="Indicator to pressure relationship across all ecosystems."}
all_stats %>%
  filter(n > 20) %>%
  ggplot(aes(x = HIF, y = w_mean))+
  geom_point(size=2, position = position_dodge2(.1))+
  geom_violin(alpha=0)+
  theme_bw()
```

There is a trend here, but not very strong.

@@

```{r, fig.cap="Sample size against indicator uncertainty."}
all_stats %>%
  filter(n > 5) %>%
  ggplot(aes(x = n, y = sd))+
  geom_point(size=2, position = position_dodge2(.1))+
  theme_bw()+
    facet_wrap(.~eco, scales="free")
```

This shows that the uncertainty is really inflated with sample sizes less than about 50-100.

```{r, fig.cap="Summary statistics for the alien species indicator."}
DT::datatable(all_stats) %>%
  formatRound(columns=3:8, digits=2)
```

@@

```{r, eval=F}
corrCheck <- st_intersection(alien_data, HIA_muni)
saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
```
